---
title: "Problem Set 4"
subtitle: "Due date: 16 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 30**

## Question 1

*Points: 5*

Suppose I am playing poker with a regular deck of 52 cards. I have three cards in my hand, one of which is a king. What is the probability that the next two cards I draw will both be kings?

There are 4 kings in a deck of 52 cards.\
P(A)=number of outcomes in event A/number of total outcomes in the sample space

P(One King) = 3 kings left after the one in your hand / 51 cards -\> 3/51 or 0.059

P(Next King) = 3 kings left after the one in your hand / 50 cards -\> 3/50 or 0.06

P(A & B)=P(A)\*P(B)

P(One King & Next King) =P(One King) \* P(Next King)

(3/51) \* (3/50)

```{r}
(3/51) * (3/50)
```

**Answer - The probability that the next two cards you draw will both kings is about .00353 or .353%**

## Question 2

*Points: 5*

A survey of 1,500 randomly selected individuals reveals that 46% of them have a college degree and that 31% of them have a gun in their house. If you were to assume that college graduation and gun ownership are independent of one another, what is the probability that an individual is both a college graduate and does not have a gun in his or her house?

The complement for individuals with a gun in the house is 69%. Meaning that in the survey of 1,500 randomly selected individuals, 69% of them do not have a gun in their house.

P(College Degree) = .31

P(No Gun) = .69

P(College Degree & No Gun) = .31 \* .69

```{r}
.31 * .69
```

**Answer - The probability that an individual is both a college graduate and does not have a gun in their house is 21.39%.**

## Question 3

*Points: 5*

I roll a fair six-sided dice and I get a number greater than three. What is the probability that the next two dice rolls will also generate numbers greater than three?

```{r}
1/2 *1/2 
```

**Answer - Since each roll of six-sided dice is independent, the outcome of one roll doesn't effect the next roll. The options greater than three are four, five, and six. With two more rolls of the dice, it is a 1/2 chance for each roll to land on four, five, or six. The probability that the next two rolls will generate numbers greater than three is 25%.**

## Question 4

*Points: 10*

Assume that in the United States, 15% of people are smokers and, of smokers, 60% smoke at least one pack a day. Among those who smoke at least a pack a day, 60% will develop lung cancer at some point in their lives while, in the rest of the population (which includes smokers who smoke less than a pack a day and non-smokers), only 10% will. What is the probability that a person who develops lung cancer was a smoker who smoked at least a pack a day?

-   P(C) = person who gets lung cancer

-   P(S) = person who is a smoker that smokes 1 pack a day = 0.15 \* 0.6 =0.90

-   P(C\|LN) = probability of getting cancer for smokers who smoke less than a pack a day and non-smokers = 0.10

-   P(S\|C) = probability that a person who is a smoker and smokes at least 1 pack a day will get lung cancer = 0.60

-   probability of that a person who develops lung cancer was a smoker who smoked at least a pack a day

    -   P(C) = P(C\|S) \* (P(S) + P(C\|LN)\*PLN

P(C) = (0.60 \* 0.15 \* 0.60) + (0.10 \* (1 - 0.15 \*0.60)

```{r}
(0.60 * 0.15 * 0.60)
0.10 * (1 - 0.15 *0.60) 
0.054 + 0.091
```

P(C) = 0.145 or 14.5%

P(S\|C) = P(C\|S) \* P(S)/P(C\|S) \* (P(S) + P(C\|LN) \* (PLN)

-   0.60 \* 0.091 / 0.60 \* 0.091 + (0.10 \* 0.91)

```{r}
0.60 *0.091
0.10 * .91
0.0546 + 0.091
```

```{r}
0.0546 / 0.14546
```

-   0.0546 / 0.1456 = 0.3754

**Answer - The probability of a person who smokes and who smokes at least one pack a day developing lung cancer is 37.54%.**

## Question 5

*Points: 5*

Assume that in the world, 60% of countries are democracies and 80% of countries have ratified the Rome Statute for the International Criminal Court. Assume further that 90% of democracies have ratified the Rome Accords. What is the probability that a country is either a democracy or has ratified the Rome Accords?

Countries that are a democracy - P(D)= 0.60

Countries that have ratified the Rome Statute for the ICC - P(RS) =0.80

Countries that have ratified the Rome Accords given that they are a democracy - P(RA\|D) = 0.90

P(D or RA) = P(D) + P(RA\|D) - P(D and RA\|D)

P(D or RA) = 0.60 + 0.90 - (0.60 \* 0.90) = 0.96 or 96%

```{r}
0.60 + 0.90 
0.60 * 0.90
1.5 - 0.54
```

**Answer - The probability that a country is either a democracy or has ratified the Rome Accords is 96%.**
