---
title: "Final Exam"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

Please read all of the questions and tables carefully and follow all instructions. Each question has an allotted point value. Be as thorough and specific as possible; extra calculations and incorrect information, even in the presence of correct information, will result in point deductions. Be sure to show all formulas and all necessary work to answer the questions. You may upload your completed exam to the Elms course site.

::: callout-note
While this is an open-note exam, you are not to receive assistance from anyone else. As usual, the Honor Code applies.
:::

**Total points: 50**

## Conceptual Questions

Please include all work (and computations) necessary to answer the questions.

**Total points: 20 (2 points each)**

### Question 1

::: {.callout-tip}
2/2
:::

Suppose you estimate a bivariate regression model with a total sample size of 50 and obtain a standard error for $\beta_1$ of 2.50. What is the smallest regression coefficient ($\beta_1$) that you could have and still be able to reject the null hypothesis (of no relationship between $X_1$ and $Y$) at the 0.05 level with a one-tailed test?

```{r}
critical_value1 <- qt(0.05, df = 48, lower.tail = FALSE) 
beta1_Q1 <- 2.50

smallest_beta1_Q1<- critical_value1 * beta1_Q1

cat("The smallest regression coefficient:", smallest_beta1_Q1, "\n")

```

**Answer -** The smallest regression coefficient for $\beta_1$ is 4.19306.

### Question 2

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and find that $\beta_1 = 4.20$, $SE_{\beta_1} = 1.90$, $t = 2.21$, $p = 0.01$, and a 95% confidence interval is \[0.48, 7.92\]. What is my best estimate of the effect of a two-unit change in $X$ on the mean of $Y$ in the population?

**Answer -** The best estimate of the effect of the two unit change in the $X_1$ on mean of $Y$ is 8.40 units.

### Question 3

::: {.callout-tip}
2/2
:::

I conduct an OLS regression with a sample size of 90 and 5 independent variables. To determine a p-value for each coefficient, I would examine a t-distribution with how many degrees of freedom?

**Answer -** You would examine a t-distribution with 84 degrees of freedom when determining the p-value for each coefficient.

### Question 4

::: {.callout-tip}
0/2
:::

Suppose I regress $Y$ on $X$ and compute the mean response for $Y$ at some specified value of $X$. When determining the confidence interval around this mean response, which of the following will **NOT** have any effect on the width (or, size) of that confidence interval?

A.  Total sample size,

B.  Mean-squared error,

C.  The specified value of $X$,

D.  *None of the above: all of these (above) will affect the confidence interval*,

E.  There is not enough information to answer this question.

**Answer -** C. The specified value of $X$

### Question 5

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X$ and find that $\beta_1$ has a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  The lower and upper bound of a 95% confidence interval around $\beta_1$ will have the same sign,

B.  The absolute value for the t-statistic for $\beta_1$ will be greater than 1.96,

C.  A 90% confidence interval around $\beta_1$ will not contain zero,

D.  All of the above,

E.  None of the above and/or there is not enough information.

**Answer -** D. All of the above

### Question 6

::: {.callout-tip}
0/2
:::

I regress $Y$ on three independent variables -- $X_1$, $X_2$, and $X_3$ -- and I find the following 95% confidence intervals -- $\beta_1$: \[0.12, 1.45\], $\beta_2$: \[-0.01, 0.15\], and $\beta_3$: \[-0.64, -0.01\]. Which of the following statements is most accurate?

A.  $\beta_2$ and $\beta_3$ will have negative coefficients, and $\beta_1$ is statistically significant at the 0.05 level (two-tailed),

B.  $\beta_1$ will have a positive coefficient, and $\beta_1$ is the only statistically significant coefficient (of the three coefficients) at the 0.05 level (two-tailed),

C.  $\beta_1$ will have a positive coefficient, $\beta_2$ and $\beta_3$ will have negative coefficients, and only $\beta_1$ and $\beta_3$ are statistically significant at the 0.05 level (two-tailed),

D.  All of the above,

E.  *None of the above and/or there is not enough information.*

**Answer -** C. $\beta_1$ will have a positive coefficient, $\beta_2$ and $\beta_3$ will have negative coefficients, and only $\beta_1$ and $\beta_3$ are statistically significant at the 0.05 level (two-tailed)

### Question 7

::: {.callout-tip}
2/2
:::

Suppose I estimate a regression with two independent variables and obtain a $R^2$ of 0.40 where the Residual Sum of Squares is equal to 5,150. What does the Total Sum of Squares equal in this regression model?

```{r}
R2_Q7 <- 0.40
RSS7 <- 5150

TSS7 <- RSS7 / (1 - R2_Q7)

print(TSS7)

```

**Answer -** The Total Sum of Squares is 8583.333

### Question 8

::: {.callout-tip}
0/2
:::

I regress $Y$ on $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$). I find that $\beta_1 = -0.87$ with a two-tailed p value of 0.001 and $\beta_2 = 3.52$ with a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  $X_2$ has a larger substantive effect on $Y$ than $X_1$, but the effect of $X_1$ is more statistically significant,

B.  $X_1$ has a larger substantive effect on $Y$ than $X_2$ and $X_1$ is more statistically significant than $X_2$,

C.  $X_1$ has a larger substantive effect on $Y$ than $X_2$, but the effect of $X_2$ is more statistically significant,

D.  *None of the above and/or there is not enough information.*

**Answer -** A. $X_2$ has a larger substantive effect on $Y$ than $X_1$, but the effect of $X_1$ is more statistically significant

### Question 9

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and obtain a $R^2$ of 0.45. Then, I regress $Y$ on both $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$) and obtain a $R^2$ of 0.53. Which of the following statements must be true?

A.  The correlation between $X_1$ and $Y$ is stronger than the correlation between $X_2$ and $Y$,

B.  The coefficient on $\beta_1$ is statistically significant in both models,

C.  The model with $X_1$ and $X_2$ explains more of the variation in $Y$ than a model with just $X_1$,

D.  All of the above: all of these statements are true.

**Answer -** C. The model with $X_1$ and $X_2$ explains more of the variation in $Y$ than a model with just $X_1$

### Question 10

::: {.callout-tip}
1/2
:::

Based on the following regression equation, might the negative coefficient on the South variable be the result of southern states having a lower proportion of high school graduates than non-southern states? In no more than one sentence, explain your answer.

$$
Turnout = 34 + 0.5(Percent\ High\ School\ Graduates) – 5.9(South)
$$

**Answer -** The negative coefficient does not provide enough information to about the southern states having a lower proportion, it only measures that the association between the southern states and voter turnout.

::: {.callout-tip}
What is this association? 
:::

## Applied Questions

All datasets referenced below are posted on the ELMs course site.

### Question 1

*Points: 15*

Use the `world` dataset to answer the following questions. The relevant variables and their coding information are as follows: `literacy` indicates a country's literacy rate; `dem_score14` represents a country's level of democratization (higher values indicate greater democratization); `spendeduc` reflects the amount of public expenditures on education as a percentage of GDP; `gdp_10_thou` represents GDP per capita (one unit is \$10,000 USD); `educ_quality` indicates the average quality rating of a country's educational system; and `ungr9095` represents the percent average annual population growth.

#### Part A

*Points: 4*

::: {.callout-tip}
3/4
:::

Evaluate the bivariate relationship (using R) between a country's support for public education and its literacy rate. Report the results in a professionally formatted table. Do countries that spend more money on education appear to have a higher literacy rate?

```{r}
library(poliscidata)

model1 <- lm(literacy ~ spendeduc, data = world)

modelsummary::modelsummary(model1,
                           statistic = c("t = {statistic}", "SE = {std.error}", "conf.int", "p.value"),
                           coef_rename = c("spendeduc" = "Public Expenditures on Education (% of GDP)"),
                           gof_map = "nobs")

library(stargazer)

stargazer(model1, type = "text",
          title = "Regression",
          covariate.labels = c("Education Expenditure (% of GDP)", "Intercept"),
          align = TRUE, 
          single.row = TRUE,
          star.cutoffs = c(0.05, 0.01, 0.001),
          digits = 3
)

summary(model1)


```

**Answer -** The regression results indicate that there is a relationship between a country's spending on education and its literacy rate. There is a positive relationship between spending on education and literacy rate. The p-value of 0.0505 suggests that the relationship is marginally significant at the 0.05 significance level. 

::: {.callout-tip}
The relationship is not statistically significant at this level. 
:::

While the coefficient is positive, indicating an increase in literacy rate with increased spending on education, the practical significance might be influenced by other factors not included in the model.

#### Part B

*Points: 5*

::: {.callout-tip}
4/5
:::

Estimate a second regression model (using R) that includes both a country's public expenditures on education and its level of democratization as independent variables. When controlling for democratization, does education expenditures exhibit a significant impact on literacy? Report the results in a professionally formatted table. Be sure to discuss the extent to which the results change, and if so, why they changed.

```{r}
model1B <- lm(literacy ~ spendeduc + dem_score14, data = world)

modelsummary::modelsummary(model1B,
                           statistic = c("t = {statistic}", "SE = {std.error}", "conf.int", "p.value"),
                           coef_rename = c("dem_score14" = "A Country's Level of Democratization"),
                           gof_map = "nobs")

stargazer(model1B, type = "text",
          title = "Regression",
          covariate.labels = c("A Country's Level of Democratization", "Intercept"),
          align = TRUE,
          single.row = TRUE,
          star.cutoffs = c(0.05, 0.01, 0.001),
          digits = 3)

summary(model1B)
```

**Answer -** The regression results indicate that, when controlling for democratization, education expenditures do not exhibit a statistically significant impact on literacy rates. The coefficient for 'spendeduc' is 0.2582. This suggests that, on average, a one-unit increase in education expenditure is associated with a 0.2582 unit increase in literacy rate. 

::: {.callout-tip}
You should use the variable units. Do not write "one-unit". 
:::

However, the p-value for this coefficient is not statistically significant (p-value = 0.748). The results suggest that education expenditure does not have a significant independent impact on literacy rates when considering the influence of democratization.

#### Part C

*Points: 6*

::: {.callout-tip}
4/6
:::

Now estimate a third model (using R) that includes all of the variables listed above. Report the regression results in a professionally formatted table and interpret each coefficient. When interpreting the impact of democratization using this regression model, present/utilize a visual representation (using R) of the expected change in literacy as a function of democratization.

```{r}
world <- read.csv("world.csv")

model1C <- lm(literacy ~ dem_score14 + spendeduc + gdp_10_thou + educ_quality + ungr9095, data = world) 

modelsummary::modelsummary(model1C,
                           statistic = c("t = {statistic}", "SE = {std.error}", "conf.int", "p.value"),
                           coef_rename = c("dem_score14" = "A Country's Level of Democratization",
                                           "spendeduc" = "Public Expenditures on Education (% on GPD)",
                                           "gdp_10_thou" = "GDP Per Captia",
                                           "educ_quality" = "Average Quality Rating of a Country's Educational System",
                                           "ungr9095" = "Annual Population Growth (Percent Average)"),
                           gof_map = "nobs")

summary(model1C)
```

```{r}
library(marginaleffects)
library(plotly)
plot_predictions(model1C, condition = "dem_score14")
```

**Answer -** The estimated literacy rate when all predictor variables are zero is 55.9103. A one-unit increase in democratization score is associated with a 2.5456 unit increase in literacy rate. 

::: {.callout-tip}
You should use the variable units. Do not write "one-unit". 
:::

This coefficient is statistically significant with a p-value of 0.006395. The coefficient for 'spendeduc' is 0.1124. This suggests that, on average, a one-unit increase in education expenditure is associated with a 0.1124 unit increase in literacy rate. 

::: {.callout-tip}
You should use the variable names.  
:::

However, the p-value for this coefficient is not statistically significant (p-value = 0.903342). The coefficient for 'gdp_10_thou' is 2.3598. This implies that, on average, a one-unit increase in GDP per capita is associated with a 2.3598 unit increase in literacy rate. However, the p-value for this coefficient is not statistically significant (p-value = 0.245593).The coefficient for 'educ_quality' is 3.8118. This suggests that, on average, a one-unit increase in education quality is associated with a 3.8118 unit increase in literacy rate. However, the p-value for this coefficient is marginally significant (p-value = 0.086568). A one-unit increase in 'ungr9095' is associated with a -3.3259 unit decrease in literacy rate. This coefficient is statistically significant with a very low p-value (p-value = 0.000626).

### Question 2

*Points: 15*

Using the `states` dataset, use R to regress the variable `obama08` (the percentage of a state's vote that President Obama received in the 2008 U.S. presidential election) on the following independent variables: `cig_tax` represents the amount of a state's cigarette tax (in dollars); `college` is the percentage of a state's population that graduated college (0 to 100); `union07` is the percentage of a state's workers that are union members (0 to 100); and `south` is a dichotomous variable coded as `1` if the state is located in the South (`0` otherwise). Report the results in a professionally formatted table. Answer the following questions.

```{r}
states <- read.csv("states.csv")

model2 <- lm(obama08 ~ cig_tax + college + union07 + south, data = states) 

modelsummary::modelsummary(model2,
                           statistic = c("t = {statistic}", "SE = {std.error}", "conf.int", "p.value"),
                           coef_rename = c("cig_tax" = "State's Cigarette Tax (in Dollars)",
                                           "college" = "% of State's Population that Graduated College (0 to 100)",
                                           "union07" = "% of State's Workers that are Union Members (0 to 100)"),
                           gof_map = "nobs") 

summary(model2)
```

#### Part A

*Points: 6*

::: {.callout-tip}
4/6
:::

Interpret the substantive results from this regression model. Be sure to discuss the direction, magnitude, and statistical significance of each slope coefficient. Does the `union07` variably exhibit a substantively significant impact on voting behavior in the 2008 election?

**Answer -**

The coefficient for 'cig_tax' is 0.5230. This suggests that, on average, a one-unit increase in cigarette taxes is associated with a 0.5230 percentage point increase in the vote share for Obama. However, the p-value for this coefficient is not statistically significant (p-value = 0.815676), indicating that this effect is not statistically reliable.

::: {.callout-tip}
You should use the variable units. Do not write "one-unit". And the variable names. 
:::

The coefficient for 'college' is 0.9646. This implies that, on average, a one-unit increase in the percentage of the population with a college degree is associated with a 0.9646 percentage point increase in the vote share for Obama. This coefficient is statistically significant with a very low p-value (p-value = 0.000299), suggesting a substantively significant impact.

The coefficient for 'union07' is 0.7514. This indicates that, on average, a one-unit increase in the percentage of union members in 2007 is associated with a 0.7514 percentage point increase in the vote share for Obama. This coefficient is statistically significant with a p-value of 0.003511, suggesting a substantively significant impact.

The coefficient for 'south' is 1.3928. This suggests that, on average, being in the southern region is associated with a 1.3928 percentage point increase in the vote share for Obama. However, the p-value for this coefficient is not statistically significant (p-value = 0.593179), indicating that this effect is not statistically reliable.

All fo the coefficients present a positive direction.

Both 'college' and 'union07' are statistically significant at conventional levels, indicating that these variables have a significant impact on the vote share for Obama with higher college-educated populations and a greater percentage of union members associated with increased support for Obama.

#### Part B

*Points: 2*

::: {.callout-tip}
2/2
:::

Interpret the intercept coefficient. What does this represent and is it meaningful to interpret this coefficient on its own? Why or why not?

**Answer -** The estimated mean value of Obama08 (percentage of votes for Obama in 2008) when all predictor variables are zero is 16.1280%. It is overall not meaningful to interpret the intercept coefficient, but if one does use the intercept the interpretation should be done cautiously because, In this model, 'south' is a binary categorical variable (indicating whether a state is in the southern region or not). When 'south' is zero (not in the southern region), the intercept is applicable. However, the interpretation of a state not being in the southern region might not have a straightforward meaning, and it's not practically meaningful to have all categorical variables at zero. It might be more relevant to interpret the intercept in conjunction with the other coefficients, especially the ones representing meaningful changes in predictor variables.

#### Part C

*Points: 2*

::: {.callout-tip}
2/2
:::

What is the expected mean percentage of Obama's vote total in a Southern state with a cigarette tax of \$2.00, a state population where 30% of people are college graduates, and where 15% of workers in a state are union members?

```{r}
intercept <- 16.1280
cig_tax_coeff <- 0.5230
college_coeff <- 0.9646
union07_coeff <- 0.7514
south_coeff <- 1.3928


cig_tax_value <- 2.00
college_value <- 30 
union07_value <- 15
south_value <- 1

expected_mean_percentage <- intercept +
  cig_tax_coeff * cig_tax_value +
  college_coeff * college_value +
  union07_coeff * union07_value +
  south_coeff * south_value


cat("Expected mean % of Obama's vote:",expected_mean_percentage, "%\n" )
```

#### Part D

*Points: 3*

::: {.callout-tip}
3/3
:::

What is a 90% confidence interval around the coefficient for the `south` variable.

```{r}
model2D <- lm(obama08 ~ cig_tax + college + union07 + south, data = states)

conf_interval2D <- confint(model2D, "south", level = 0.90)

cat("90% Confidence Interval for the 'south' variable:\n", conf_interval2D, "\n")

```

#### Part E

*Points: 3*

::: {.callout-tip}
3/3
:::

What is the null hypothesis that the F-test in this regression output is testing?

-   Null hypothesis: cig_tax = 0, college = 0, union7 = 0, south = 0

-   Independent variables = 0

-   Alternative hypothesis: One or more of these coefficients is not equal to zero.

-   F-test: 12.19 on 4 and 45 degrees of freedom

**Answer -** The F-test assesses whether there is a significant linear relationship between the dependent variable (obama08) and all the independent variables in the model combined. The null hypothesis being evaluated is that the coefficients of all independent variables are equal to zero. A low p-value from the F-statistic indicates rejection of the null hypothesis, suggesting that at least one independent variable is a predictor of the dependent variable (obama08).
