---
title: "Mid-term Exam #2"
subtitle: "Due date: 10 November"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

Please read all of the questions carefully and follow all instructions. Each question has an allotted point value. Be as thorough and specific as possible; extra calculations and incorrect information, even in the presence of correct information, will result in point deductions. Be sure to show all formulas and all necessary work to answer the questions. You may upload your completed exam to the Elms course site (attach to Midterm Exam #2).

::: callout-note
While this is an open-note exam, you are not to receive assistance from anyone else (as usual, the Honor Code applies).
:::

**Total points: 50 points**

## Conceptual questions

::: callout-note
Please include all work (and computations) necessary to answer the questions.
:::

**Total points: 26**

### Question 1

*2 points*

::: {.callout-tip}
1/2
:::

Suppose I am interested in determining if freshman undergraduates at the University of Maryland spend more hours studying in the average week than sophomore undergraduates. I conduct a study in which I take a simple random sample (SRS) of 1200 freshman students and 1200 sophomore students. I find that the freshman students in my sample study for, on average, 412 minutes per week and the sophomore students in my sample study for, on average, 335 minutes per week. The standard error of the difference is 30. What is a 90% confidence interval for the difference between freshman and sophomore students?

```{r}
fresh_mean <- 412  # average study time for freshman
soph_mean <- 335  # average study time for sophomore
se1 <- 30  # standard error

confidence_level1 <- 0.90

z1 <- qnorm(1 - (1 - confidence_level1) / 2)

difference1 <- fresh_mean - soph_mean
margin_of_error1 <- z1 * se1

lower_bound1 <- difference1 - margin_of_error1
upper_bound1 <- difference1 + margin_of_error1

cat("90% Confidence Interval:", lower_bound1, "to", upper_bound1, "minutes\n")

```

::: {.callout-tip}
You need to use the t-statistic, not z score, when working with means. 
:::

**Answer - The 90% confidence interval for the difference between freshman and sophomore students is 27.654 to 126.347 minutes.**

### Question 2

*2 points*

::: {.callout-tip}
2/2
:::

Based on the results of my study described in question 1, can you reject the null hypothesis of no statistically meaningful difference in the study habits of sophomore and freshman students? Why or why not?

**Answer - Since the 90% confidence interval does not include 0, I am pretty confident that there is a statistically significant difference between the study habits of freshman and sophomore students. Meaning, it is reasonable to conclude that freshman students, on average, spend more time studying per week compared to sophomore students.**

### Question 3

*2 points*

::: {.callout-tip}
0/2
:::

If I am testing a null hypothesis that X has no effect on Y in the population (and thus my alternative hypothesis is that X does have an effect), why might I prefer to commit a Type-II error over a Type-I error (and, of course, this holds aside my first preference of making no error at all)? Answer in no more than two sentences.

**Answer - If you are testing a null hypothesis that X has no effect on Y, preferring to commit a Type-II error over a Type-I error would be better. This decision would depend on the specific context and the relative importance of avoiding each type of error.**

::: {.callout-tip}
I need your thoughts on this. What contexts?
:::

### Question 4

*1 point*

::: {.callout-tip}
1/1
:::

When conducting a difference-of-means test, which of the following samples would yield a sampling distribution with the lowest variability?

A.  A sample of 900 with a standard deviation of 15
B.  A sample of 25 with a standard deviation of 10
C.  A sample of 625 with a standard deviation of 20
D.  A sample of 100 with a standard deviation of 6

**Answer - A. A sample of 900 with a standard deviation of 15.**

### Question 5

*1 point*

::: {.callout-tip}
1/1
:::

Which of the following probabilities is not independent?

A.  The probability that the roulette wheel will end up on 23 two times in a row.
B.  The probability that three successive coin tosses will each turn up heads.
C.  The probability that I draw an ace and then a king in a row from a deck of cards (when drawing a two-card hand).
D.  The probability that I will get a 6 on three subsequent die rolls.
E.  None of the above -- all are independent probabilities.

**Answer - C. The probability that I draw an ace and then a king in a row from a deck of cards (when drawing a two-card hand).**

### Question 6

*1 point*

::: {.callout-tip}
1/1
:::

I conduct a two-tailed difference-of-means test and obtain a t-statistic of 2.10. Is my result statistically significant (with 95% confidence)?

A.  Yes, at the 0.05 level (p\<0.05)
B.  Yes, but only at the 0.10 level (p\<0.10)
C.  Not at either the 0.05 or the 0.10 level
D.  There is not enough information to answer this question.

**Answer - D. There is not enough information to answer this question. There's no degrees of freedom nor a critical t-test.**

### Question 7

*1 point*

::: {.callout-tip}
1/1
:::

I take a sample of 1800 adults and find that 360 of them watched last Monday's NFL game. What probability represents the complement to the sample proportion of adults who watched the NFL game?

*The compliment to this probability is* 1 - *p*. 1 - 0.2 = 0.8

So, the probability representing the complement to the sample proportion of adults who watched the NFL game is 0.8 or 80%.

**Answer - The probability representing the complement to the sample proportion of adults who watched the NFL game is 0.8 or 80%.**

### Question 8

*1 point*

::: {.callout-tip}
1/1
:::

Which of the following makes it more likely that a given sample statistic will be statistically different from zero (and thus allow you to reject the null hypothesis, all else equal)?

A.  More observations
B.  Greater variance in the sample
C.  Using a two-tailed instead of a one-tailed significance test
D.  A larger confidence interval around the test statistic
E.  Both (a) and (b) (but not (c) or (d))
F.  All of the above increase the chances of statistical significance
G.  None of the above

**Answer - E. Both (a) and (b) (but not (c) or (d)).**

### Question 9

*1 point*

::: {.callout-tip}
0/1
:::

The p-value for a two-tailed test of the null hypothesis $H_0: \mu=40$ is 0.06. Which of the following statements is accurate?

A.  A 95% confidence interval around the sample mean includes the value 40
B.  A 90% confidence interval around the sample mean includes the value 40
C.  A 99% confidence interval around the sample mean does not include the value 40
D.  None of the above statements are correct
E.  All of the above statements are correct

**Answer - D. None of the above statements are correct.**

::: {.callout-tip}
The answer is A. 
:::

### Question 10

*4 points total*

#### Part A

*2 points*

::: {.callout-tip}
1/2
:::

If you roll a fair (six-sided) die twice, what is the probability that both rolls will be odd or greater than four?

$$
\frac{1}{2}*\frac{1}{2} = \frac{1}{4}
$$

$$
\frac{1}{3}*\frac{1}{3} = \frac{1}{9}
$$

$$
\frac{1}{9}+\frac{1}{4} = \frac{13}{36}
$$

**Answer - There are 3 odd numbers out of 6 possible outcomes, so the probability of rolling an odd number on one die is 3/6 or 1/2. There are 2 numbers greater than four out of 6 possible outcomes, so the probability of rolling a number greater than four on one die is 2/6 or 1/3. The probability that both rolls will be odd or greater than four is 13/36 or 0.36 or 36%.**

::: {.callout-tip}
You need to account for the intersection ({5,5}).
:::

#### Part B

*2 points*

::: {.callout-tip}
2/2
:::

The following is a distribution of U.S. college students classified by their age and full- vs. part-time status. Based on these data, what is the probability that a student is in the 25-29 age group, given that (i.e., conditional on knowledge that) the student is full time?

```{r}
tibble::tibble(
  age = c("15 - 19", "20 - 24", "25 - 29", "30+"),
  full_time = c(155, 255, 75, 35),
  part_time = c(20, 55, 80, 95)
) |> 
  knitr::kable()
```

$$
P(A|B) = \frac{P (A \cap\ B)}{P(B)}
$$

$$
P(A|B) = \frac{75}{520}
$$

**Answer - The probability that a student is in the age group 25-29 given that the student is full-time is 14.42%.**

### Question 11

*4 points total*

#### Part A

*2 points*

::: {.callout-tip}
1/2
:::

Using a SRS of 1211 people, I estimate that the proportion of people living in the South that favor teaching sexual education in public schools is 0.874 and the proportion of people outside of the south that favor teaching sexual education in public schools is 0.906. And, the standard error of the difference (in citizen views about teaching sexual education in public schools) between people living in the south and those not living in the south is 0.015. Give an interval estimate for the difference in the proportion of people favoring sex education in public schools between people who do, and do not, live in the south.

```{r}
south_proportion <- 0.874
not_south_proportion <- 0.906
sample_size11 <- 1211
se_difference <- 0.015

se11 <- sqrt((south_proportion * (1 - south_proportion) / not_south_proportion) + (not_south_proportion * (1 - not_south_proportion) / sample_size11))

z11 <- qnorm(1 - 0.05 / 2)
margin_of_error11 <- z11 * se11

difference11 <- south_proportion - not_south_proportion
lower_bound11 <- difference11 - margin_of_error11
upper_bound11 <- difference11 + margin_of_error11

cat("The 95% confidence interval for the difference in proportions is:", lower_bound11, "to", upper_bound11)

```

::: {.callout-tip}
You already have the SE of the difference: 0.015.
:::

**Answer - The 95% confidence interval of the difference in proportions is -0.715519 to 0.651519.**

#### Part B

*2 points*

::: {.callout-tip}
2/2 (error carried forward)
:::

Do the data (i.e., estimates above) show support for my hypothesis? How do you know?

**Answer - Since this confidence interval includes 0, it means that the difference in proportions could be within the range of the confidence interval. This suggests that we fail to reject the null hypothesis, which would state that there is no difference in the proportions of people favoring teaching sexual education in public schools between those living in the South and those living outside of the South.**

### Question 12

*6 points total*

I am interested in estimating the average number of texts that University of Maryland undergraduate students send in a day. My hypothesis is that the average number of texts students send is greater than 100. I take a SRS of 1600 students and find that the mean number of texts they send is 105, and with a standard deviation of 120.

#### Part A

*2 points*

::: {.callout-tip}
2/2
:::

What is a 95% confidence interval around the sample statistic?

```{r}
sample_size12 <- 1600
mean12 <- 105
sd12 <- 120

se12 <- sd12 / sqrt(sample_size12)

margin_of_error12 <- 1.96 * se12

confidence_interval11 <- c(mean12 - margin_of_error12, mean12 + margin_of_error12)

cat("The 95% confidence interval for the sample statistic is:", 99.12, "to", 110.88)
```

**Answer - The 95% confidence interval for the sample statistic is 99.12 to 110.88.**

#### Part B

*2 points*

::: {.callout-tip}
2/2
:::

When testing the null hypothesis, what is the test statistic associated with the sample statistic?

$$
 t = \frac{\bar{x} - \mu_0}{\frac{s}{\sqrt{n}}}
$$

$$
t = \frac{105 - 100}{\frac{120}{\sqrt{1600}}}
$$

**Answer - To test the null hypothesis, you can use a t-test for a single sample mean. The t statistic associated with the sample statistic is *t* = 1.67.**

#### Part C

*2 points*

::: {.callout-tip}
2/2
:::

If using a one-tailed test of the null hypothesis and you are willing to accept a Type-I error rate of 0.05, do the data support my hypothesis? Why or why not?

```{r}
library(stats)

p_value <- 1 - pt(1.67, 1599)
p_value
```

**Answer - The data supports the hypothesis that the average number of texts is greater than 100 since the p-value is less than the confidence level of 0.05.**

## Applied questions

::: callout-note
All datasets referenced below are available through the course elms page. Do note use the `poliscidata` package for these questions, as the specific variables referenced are not all included in that package.
:::

**Total points: 24**

```{r}
gss <- read.csv("gss.csv")
nes <- read.csv("nes.csv")
```

### Question 1

*8 points total*

I hypothesize that, among only those that were eligible to vote, people with greater confidence in the U.S. military were more likely to turnout to vote in the 2012 presidential election. Use data from the General Social Survey (i.e., the `gss` dataset) to test my hypothesis (the unit of analysis is the individual survey respondent). Specifically, use the following variables: `conarmy` (1 = a "great deal" of confidence; 2 = "only some" confidence; 3 = "hardly any"); and `vote12` (1 = voted; 2 = did not vote; 3 = ineligible). Answer the following questions.

```{r}
library(tidyr)
library(dplyr)

gss <- gss |>
  select(vote12,conarmy) |>
  drop_na()
```

#### Part A

*3 points*

::: {.callout-tip}
3/3
:::

Complete a cross-tab and interpret the results. Do the data support my hypothesis? Be sure to explain the nature of the relationship (or lack thereof, if relevant).

```{r}
library(modelsummary)
datasummary_crosstab(vote12 ~ conarmy, data = gss)
```

**Answer - The data supports the hypothesis that among eligible voters, individuals with higher confidence in the U.S. military were more likely to participate in the 2012 presidential election. Specifically, among those eligible to vote and who exercised that right, 52.8% of individuals with a great deal of confidence in the military cast their votes. In contrast, only 41% of those with some confidence and 6.2% with no confidence in the military voted. Additionally, it's noteworthy that although 52.8% of those with a great deal of confidence voted, a substantial 51.8% of individuals with the same high confidence level chose not to participate in the election.**

#### Part B

*3 points*

::: {.callout-tip}
2/3
:::

Compute (by hand) the chi-square statistic to test the null hypothesis of no relationship between these two variables. Be sure to show your work.

+------------+---------------------+---------------------+---------------------+---------------------+
| Vote       | Confidence Level    | 2                   | 3                   | Total               |
|            |                     |                     |                     |                     |
|            | 1                   |                     |                     |                     |
+============+=====================+=====================+=====================+=====================+
| **1**      | 642 -\> 52.8%       | 498 -\> 41.0%       | 75 -\> 6.2%         | 1215                |
|            |                     |                     |                     |                     |
|            | *expected:* 638.9   | *expected:* 483.8   | *expected:* 92.2785 | *x\^2:* 3.667102854 |
|            |                     |                     |                     |                     |
|            | *x\^2:* 0.0150415   | *x\^2:* 0.4167837   | *x\^2:* 3.235277581 |                     |
+------------+---------------------+---------------------+---------------------+---------------------+
| **2**      | 278 -\> 51.8%       | 208 -\> 38.7%       | 51 -\> 9.5%         | 537                 |
|            |                     |                     |                     |                     |
|            | *expected:* 282.378 | *expected:* 213.837 | *expected:* 40.7848 | *x\^2:* 2.785765166 |
|            |                     |                     |                     |                     |
|            | *x\^2:* 0.06787669  | *x\^2:* 0.159329625 | *x\^2:* 2.558558851 |                     |
+------------+---------------------+---------------------+---------------------+---------------------+
| **3**      | 77 -\> 53.5%        | 49 -\> 34.0%        | 18 -\> 12.5%        | 144                 |
|            |                     |                     |                     |                     |
|            | *expected:* 75.7215 | *expected:* 57.3418 | *expected:* 10.9367 | *x\^2:* 5.796834097 |
|            |                     |                     |                     |                     |
|            | *x\^2:* 0.021586501 | *x\^2:* 1.213523594 | *x\^2:* 4.561724002 |                     |
+------------+---------------------+---------------------+---------------------+---------------------+
| **Total**  | 997 -\> 52.6%       | 755 -\> 39.8%       | 144 -\> 7.6%        | 1896                |
|            |                     |                     |                     |                     |
|            |                     |                     |                     | *chisq:* 12.2212089 |
+------------+---------------------+---------------------+---------------------+---------------------+

::: {.callout-tip}
You needed to exclude those who were ineligible to vote. 
:::

#### Part C

*2 points*

::: {.callout-tip}
2/2 (error carried forward)
:::

Using the chi-square statistic that you computed in question 1(b), can you reject the null hypothesis of no relationship between these two variables with 95% confidence? Why, or why not?

**Answer - With a chi-squared statistic of 12.22 and 4 degrees of freedom, the critical value for rejection is 9.49 with 95% confidence. Since 12.22 is greater than 9.49, it rejects the null hypothesis at the 0.05/95% confidence level.**

### Question 2

*8 points total*

I hypothesize that citizens who do not support fracking are less conservative than those who do support fracking. Use data from the `nes` dataset to test my hypothesis (the unit of analysis is the individual survey respondent). Specifically, use the following variables: `fracking` (1 = "approve" of fracking; 2 = "middle"; 3 = "disapprove"); and `libcon7` (higher values represent less liberalism, or more conservatism). Answer the following questions.

#### Part A

*2 points*

::: {.callout-tip}
2/2
:::

Using these data, what is the point estimate for the mean conservatism/liberalism score among those that disapprove of fracking? What is the point estimate for the mean conservatism/liberalism score among those that approve of fracking?

```{r}
nes <- nes |>
  select(fracking, libcon7) |>
  drop_na()

datasummary_crosstab(libcon7 ~ fracking, data = nes)
```

```{r}
library(dplyr)

disapprove2 <- nes |>
  filter(fracking == 3) |>
   summarize(disapprove = mean(libcon7))

approve2 <- nes |>
  filter(fracking == 1) |>
  summarize(approve = mean(libcon7))

disapprove2
approve2
```

**Answer - The point estimate for the mean conservatism/liberalism score among those who disapprove of fracking is 3.5. The point estimate for the mean conservatism/liberalism score among those that approve of fracking is 5.1.**

#### Part B

*6 points*

::: {.callout-tip}
4/6
:::

Evaluate the null hypothesis that there is no difference in the mean conservatism/liberalism score among those that approve vs. disapprove of fracking. Do the data support my hypothesis? Why or why not? Be sure to show all work necessary to answer the question by hand (i.e., you may only use R to the extent that is absolutely necessary to complete the question; otherwise, you must show how you would answer the question by hand).

```{r}
mean_difference <- 3.54943 - 5.109453

standevAPR <- nes |>
  filter(fracking == 1) |>
  summarize(approve_standev2 = sd(libcon7))
#standard deviation for approve is 1.27450

standevDIS <- nes |>
  filter(fracking == 3) |>
  summarize(disapprove_standev2 = sd(libcon7))
#standard deviation for disapprove is 1.49118

sample_sizeAPR <- nes |>
  filter(fracking == 1) |>
  summarize(approve_sample2 = length(libcon7))
#sample size for approve is 804

sample_sizeDIS <- nes |>
  filter(fracking == 3) |>
  summarize(disapprove_sample3 = length(libcon7))
#sample size for disapprove is 1578

#combinded standard deviation from online equation
combinded_sd2<- sqrt(((1578 - 1) * (1.49118^2) + (804 - 1) * (1.274507^2)) / ((1578 + 804) -2))
combinded_sd2
#combinded standard deviation is 1.421771
```

::: {.callout-tip}
You needed to calculate the pooled SE. 
:::

```{r}
t2B <- mean_difference/(combinded_sd2) * (sqrt(1/1578 + 1/804))
t2B
#t statistic is -0.04754344

df2B <- 804 + 1578 -2
df2B
#degree of freedom is 2380

p_value2B <- 2 * (1 - pt(abs(t2B), df2B))
p_value2B
#p
```

**Answer - Since the p-value is greater than the confidence level of 0.05, it fails to reject the null hypothesis. Meaning, there is no significant difference in the means of those who approve and those who disapprove of fracking through the conservatism/liberalism score.**

### Question 3

*8 points total*

I hypothesize that people who express that religion is important to them were more likely to turnout to vote in the 2016 presidential election. Use data from the `nes` dataset to test my hypothesis (the unit of analysis is the individual survey respondent). Specifically, use the following variables: `Relig_imp` (0 = not important; 1 = somewhat important; 2 = quite a bit; 3 = a great deal); and `Voted_2016` (0 = did not vote; 1= voted). Answer the following questions.

```{r}
nes2 <- read.csv("nes.csv")

nes2 <- nes2 |>
  select(Relig_imp, Voted_2016) |>
  drop_na()

datasummary_crosstab(Voted_2016 ~ Relig_imp, data = nes2)
```

#### Part A

*2 points*

::: {.callout-tip}
2/2
:::

Using these data, what is the point estimate for the proportion of respondents that voted (i.e., turnout rate), among citizens expressing that religion is not important? What is the point estimate for the proportion of respondents that voted, among citizens expressing that religion matters a great deal?

```{r}
voted_not_imp_religion <- 971
total_sample_not_imp_religion <- 1268

point_estimate_voted_norel <- voted_not_imp_religion/total_sample_not_imp_religion

point_estimate_voted_norel
```

**Answer - The point estimate for the proportion of respondents that voted among citizens expressing that religion is not important is 0.7657729.**

```{r}
voted_great_deal_religion <- 995
total_sample_greatrel <- 1229

point_estimate_voted_greatrel <- voted_great_deal_religion/total_sample_greatrel

point_estimate_voted_greatrel
```

**Answer - The point estimate for the proportion of respondents that voted among citizens expressing that religion matters a great deal is 0.8096013.**

#### Part B

*6 points*

::: {.callout-tip}
2/6
:::

::: {.callout-tip}
This is a difference of proportions problem, not a difference of counts. 
:::

Evaluate the null hypothesis that there is no difference in the proportion of voters (i.e., turnout rate) among citizens expressing that religion is not important vs. those reporting that religion matters a great deal. Do the data support my hypothesis? Why or why not? Be sure to show all work necessary to answer the question by hand (i.e., you may only use R to the extent that is absolutely necessary to complete the question; otherwise, you must show how you would answer the question by hand).

+--------------+------------------+------------------+------------------+------------------+------------------+
| Turnout Rate | Religion         | 1                | 2                | 3                | Total            |
|              |                  |                  |                  |                  |                  |
|              | 0                |                  |                  |                  |                  |
+==============+==================+==================+==================+==================+==================+
| **0**        | 297              | 106              | 141              | 234              | 778              |
|              |                  |                  |                  |                  |                  |
|              | *ex:* 272.51492  | *ex:* 101.655801 | *ex:* 139.696133 | *ex:* 264.133149 | *x\^2:* 5.835584 |
|              |                  |                  |                  |                  |                  |
|              | *x\^2:* 2.200085 | *x\^2:* 0.185647 | *x\^2:* 0.012169 | *x\^2:* 3.437685 |                  |
+--------------+------------------+------------------+------------------+------------------+------------------+
| **1**        | 971              | 367              | 509              | 995              | 2842             |
|              |                  |                  |                  |                  |                  |
|              | *ex:* 995.485083 | *ex:* 371.344199 | *ex:* 510.303867 | *ex:* 964.866851 | *x\^2:* 1.597459 |
|              |                  |                  |                  |                  |                  |
|              | *x\^2:* 0.602238 | *x\^2:* 0.050821 | *x\^2:* 0.003332 | *x\^2:* 0.941069 |                  |
+--------------+------------------+------------------+------------------+------------------+------------------+
| **Total**    | 1268             | 473              | 650              | 1229             | 3620             |
|              |                  |                  |                  |                  |                  |
|              |                  |                  |                  |                  | chisq 7.43305    |
+--------------+------------------+------------------+------------------+------------------+------------------+

**Answer - With a chi squared statistic of 7.433 and 3 degrees of freedom, the critical value for rejection is 7.81 with 95% confidence. Since 7.433 is less than 7.81, it fails to reject the null hypothesis. This suggest that there is a difference in the the proportion of voters among citizens expressing that religion matters a great deal versus those expressing that religion is not important.**

**Your hypothesis that** **people who express that religion is important to them were more likely to turnout to vote in the 2016 presidential election is supported by data in the cross tab. Of those that expressed that religion matters a great deal, 35% turned out to vote as opposed to 30.1% who did not turn out to vote. I am curious to see why that trend was followed with those who expressed that religion was not important, was somewhat important, or mattered quite a bit.**
