---
title: "Problem Set 5"
subtitle: "Due date: 23 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 25**

```{r}
library(scales)
library(tidyverse)
```

## Question 1

*Total points: 6*

::: {.callout-tip}
4/6
:::

Use the data in the table below to answer the following questions.

```{r}
#| echo: false

tibble::tibble(
  `Age group` = c("18-24", "25-30", "31 and up", "TOTAL"),
  `Non-voters` = c(70, 40, 220, 330),
  Voters = c(50, 50, 570, 670),
  Total = `Non-voters` + Voters
) |> 
  knitr::kable(caption = "Voting by Age in 2000")
```

### Part A

*Points: 2*

::: {.callout-tip}
2/2
:::

What is the probability of being 25-30 or a non-voter?

P(A or B) = P(A) + P(B) - P(A and B)

0.09 + 0.33 - (0.04)

```{r}
0.09 + 0.33

0.42-0.04
```

**Answer - The probability of being 25-30 or a non-voter is 38%.**

### Part B

*Points: 4*

::: {.callout-tip}
2/4
:::

Assuming a normal distribution, report the 95% confidence intervals for the percentage of 18-to-24-year-olds who did not vote, and then the percentage of 25-to-30-year-olds who did not vote.

```{r}
#Confidence interval for 18-24
confidenceI <- function(p, n) {
  se <- sqrt(p*(1-p)/n)
  lower_range <- p - 1.96 * se
  upper_range <- p + 1.96 * se
  return(c(lower_range, upper_range))
}

per18to24 <- 70/120
con18to24 <- confidenceI(per18to24, 120)

#Confidence interval for 25-30
per25to30 <- 40/90
con25to30 <- confidenceI(per25to30, 90)

list(con18to24 = con18to24, con25to30 = con25to30)
```

**Answer - The confidence interval of 18 to 24 year olds who did not vote is 0.495 to 0.672. The percentage of 25 to 30 year olds who did not vote is between 34.18% to 54.71%.**

::: {.callout-tip} 
The probability that an individual is 18-to-24 and a non-voter is not conditional. Therefore, you look at the proportion of the whole population who meet those criteria.

$$
\hat{p} = Pr(18\ to\ 24 \& NV) = \frac{70}{1000} = 0.07 
$$

And: 

$$
n = 1000
$$

Therefore: 

$$
SE(\hat{p}) = \sqrt{\frac{0.07 * 0.93}{1000}} = 0.008068457
$$

The 95% confidence interval can be found:

$$
\hat{p} \pm 1.96*SE(\hat{p})
$$

Therefore, the 95% confidence interval for the percentage of 18-to-24-year-olds who did not vote is `r percent(0.07 - 1.96 * 0.008068457, accuracy = 0.01)` to `r percent(0.07 + 1.96 * 0.008068457, accuracy = 0.01)`.

Using the same method, the 95% confidence intervals for the percentage of 25-to-30-year-olds who did not vote is `r percent(0.04 - 1.96*0.006196773, accuracy = 0.01)` to `r percent(0.04 + 1.96*0.006196773, accuracy = 0.01)`.
:::

## Question 2

*Total points:* *7*

Assume that the standard deviation for the population distribution of a state in which you want to conduct a poll is 200.

### Part A

*Points: 3*

::: {.callout-tip}
3/3
:::

Calculate the spread of the sampling distribution for each of the following sample sizes: 1, 4, 25, 100, 250, 1000, 5,000, and 10,000.

```{r}
stanDeviation <- 200

sample_distribution <- c(1, 4, 25, 100, 250, 1000, 5000, 10000) 

se <- stanDeviation / sqrt(sample_distribution)

data.frame(sample_distribution, standard_error = se)
```

### Part B

*Points: 1*

::: {.callout-tip}
1/1
:::

Describe specifically how the variability of the sampling distribution changes as the sample size varies. Considering the expense of running a poll, which sample size do you think is most optimal if conducting the poll?

**Answer - As the sampling size increases, the variability of the standard deviation decreases. This implies that the standard error and sample distribution has an inverse relationship. Considering the expense of running a poll, I would use a sample size of 1000 since larger samples tend to yield more accurate results.**

### Part C

*Points: 3*

::: {.callout-tip}
3/3
:::

Display your results graphically (using R) with the sample size on the x-axis and the standard error (of the sampling distribution) on the y-axis.

```{r}

df <- data.frame(sample_distribution, standard_error = se)

library(ggplot2)

ggplot(data = df, mapping = aes(x=sample_distribution, y = standard_error)) + 
  geom_point(color = "blue", size = 3)+
  geom_line(color = "black", size = 1) +
  scale_x_log10(labels = scales::comma_format()) + 
  labs(
    x = "Sample Size",
    y = "Standard Error"
  )
```

## Question 3

*Points: 4*

::: {.callout-tip}
4/4
:::

Suppose you conduct a survey (to generate a sample mean of interest) and find that it has a margin of error of 4.5 with a sample size of 900 using a 95% confidence interval. What would the margin of error be for a 90% confidence interval?

```{r}
z_95 <- 1.96
z_90 <- 1.645
ME95 <- 4.5

se3 <- ME95/z_95

se3

ME90 <- z_90 * se3

ME90
```

**Answer - The margin of error for 90% is 3.776786**

## Question 4

*Points: 4*

::: {.callout-tip}
4/4
:::

Assume that, in State A, the mean income in the population is \$20,000 with a standard deviation of \$2,000. If you took an SRS of 900 individuals from that population, what is the probability that you would get a sample mean income of \$20,200 or greater? What would be the probability if the sample size was only 25?

::: callout-note
Assume a normal distribution for both questions.
:::

```{r}
sample_mean4 <- 20200
population_mean4 <- 20000
population_std_dev4 <- 2000
sample_size1 <- 900
sample_size2 <- 25

#Probability of 900
z_score1 <- (sample_mean4 - population_mean4) / (population_std_dev4 / sqrt(sample_size1))

z_score1

probability1 <- 1 - pnorm(z_score1)

probability1

#Probability of 25
z_score2 <- (sample_mean4 - population_mean4) / (population_std_dev4 / sqrt(sample_size2))

z_score2

probability2 <- 1 - pnorm(z_score2)

probability2
```

**Answer - The probability that 900 individuals would get a sample mean income of \$20,200 is 0.13%, and the probability that 25 individuals would get a sample mean income of \$20,200 is 31%.**

## Question 5

*Points: 4*

::: {.callout-tip}
2/4
:::

Assume that a coin is fair. If I flip a coin 500 times, what is a 95% confidence interval for the range of the count of heads that I will get? What if I flip the coin 5,000 times? What about 50,000 times?

```{r}
n500 <- 500
p_hat <- 0.5
z_score5 <- qnorm(0.975)

CI_500 <- c(p_hat - z_score5 * sqrt(p_hat * (1 - p_hat) / n500),
            p_hat + z_score5 * sqrt(p_hat * (1 - p_hat) / n500))

CI_500
```

```{r}
n5000 <- 5000

CI_5000 <- c(p_hat - z_score5 * sqrt(p_hat * (1 - p_hat) / n5000),
            p_hat + z_score5 * sqrt(p_hat * (1 - p_hat) / n5000))

CI_5000
```

```{r}
n50000 <- 50000

CI_50000 <- c(p_hat - z_score5 * sqrt(p_hat * (1 - p_hat) / n50000),
            p_hat + z_score5 * sqrt(p_hat * (1 - p_hat) / n50000))

CI_50000
```

**Answer - The confidence interval for 500 flips is 0.456 to 0.544. The confidence interval for 5000 flips is 0.486 to 0.514. The confidence interval for 50000 is 0.496 to 0.504.**

::: {.callout-tip}
Almost! The question asks for the confidence interval for the range of the *count* of heads. So, your center point should the number of heads you flipped in these trials. If you have a 50% chance of getting heads when you flip a fair coin, your expected number of heads should be 250, 2,500, and 25,000 heads from each trial. You calculate your confidence around this result. 

Also, remember that the standard deviation of a binomial distribution is: 

$$
\sigma_X = \sqrt{np(1-p)}
$$

```{r}
p_head <- 0.5

tibble(
  flips = c(500, 5000, 50000),
  expected_n_heads = flips*p_head,
  sd = sqrt(flips*p_head*(1-p_head)),
  lower_ci = expected_n_heads - qnorm(0.975)*sd,
  upper_ci = expected_n_heads + qnorm(0.975)*sd
) |> 
  knitr::kable()
```
:::
