---
title: "Problem Set 1"
subtitle: "Due date: 18 September"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 25**

::: {.callout-tip title="HG comments"}
Good work. Make sure to read each question carefully and provide full answers. 18/25. 
:::

::: {.callout-tip title="HG comments"}
Make sure that you can render you .qmd file without getting an error. You need to include `library(poliscidata)` in your libraries code chunk or call it explicitly (`poliscidata::gss`) every time you use it. 

```{r}
library(poliscidata)
```
:::

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## Question 1

*Points: 5*

::: {.callout-tip title="HG comments"}
2/5.
:::

Using the `gss` dataset (which has survey data with the individual respondent as the unit of analysis), create a frequency distribution and bar chart for each of the following variables: `degree` and `partyid_3`. Describe the distribution of each variable in detail.

::: callout-note
The `gss` dataset can be found in `poliscidata::gss`. Take a look at `?gss` to see what these variables measure.
:::

**Answer -** Degree Frequency Distribution and Bar Chart

```{r}
degree_data <- select(poliscidata::gss, degree)

degree <- gss |> 
  group_by(degree) |>
  summarize(count = n())

ggplot(degree, aes(x = degree, y = count)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Degree Frequency Distribution", x = "Degree", y = "Frequency") +
  theme_minimal()

summary(degree)
```

::: {.callout-tip title="HG comments"}
You're on the right track, but this is not a frequency distribution. It's the five number summary. You can use `janitor::tabyl()` or `freqC()` or `table()` in base R to get the frequency of each category.  
:::

::: {.callout-tip title="HG comments"}
The question also asks you to describe the distribution of this variable. A quick reference to the most and least frequent variable and whether the count of the variables are very different from each other is perfect. 
:::

**Answer -** Party Identification Frequency Distribution and Bar Chart

```{r}
partyid_data <- select(poliscidata::gss, partyid_3)

partyid <- gss |> 
  group_by(partyid_3) |>
  summarize(count = n())

ggplot(partyid, aes(x = partyid_3, y = count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(title = "Party ID Frequency Distribution", x = "Party ID", y = "Frequency") +
  theme_minimal()

summary(partyid)
```

::: {.callout-tip title="HG comments"}
See comments above. 
:::

## Question 2

Examine the following vector of (fake) student IQ test scores:

```{r}
iq <- c(
  145, 139, 126, 122, 125, 130, 96, 110, 118, 118, 101, 142, 134, 124, 112, 109, 
  134, 113, 81, 113, 123, 94, 100, 136, 109, 131, 117, 110, 127, 124, 106, 124, 
  115, 133, 116, 102, 127, 117, 109, 137, 117, 90, 103, 114, 139, 101, 122, 105, 
  97, 89, 102, 108, 110, 128, 114, 112, 114, 102, 82, 101
)
```

### Part A

*Points: 5*

::: {.callout-tip title="HG comments"}
5/5.
:::

Find the five-number summary, mean, and standard deviation for these data. Also, are there any suspected outliers in the distribution? If so, what are they and how do you know?

**Answer -** Five Number Summary

```{r}
summary(iq)
```

**Answer -** Mean

```{r}
mean(iq)
```

**Answer -** Standard Deviation

```{r}
sd(iq)
```

**Answer -** Based on the box plot, there does not seem to be suspected outlier. There is no plot beyond the whiskers that would indicate an outlier is present. Also, when specifically asking R to display an outlier in the distribution, it answers with "numeric(0)".

```{r}
iq.df <- data.frame(iq)

ggplot(iq.df, aes(x = "", y = iq)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "IQ Boxplot", y = "IQ Scores") +
  theme_minimal()

boxplot.stats(iq)$out

outlier <- boxplot.stats(iq)$out
```

### Part B

*Points: 5*

::: {.callout-tip title="HG comments"}
3/5.
:::

In large populations, IQ scores are standardized to have a mean of 100 and a standard deviation of 15. In what way does the distribution among these students differ from the overall population?

**Answer -** The distribution among these students differs from the overall population, because their IQ was slightly higher than the overall population. This could be due to many factors such as testing conditions or the population of students.

::: {.callout-tip title="HG comments"}
You just need to be a bit more specific with your language. The mean IQ of the student group is higher than that of a large population. We also needed your comparison of the two SDs. They are roughly equal, meaning that these two groups are equivalently spread out around their means. 
:::

```{r}
mean <- mean(iq)

sd <- sd(iq)

standard <- (iq - mean) / sd * 15 + 100

ggplot(iq.df, aes(x = iq)) +
  geom_density(aes(color = "IQ Distribution"), fill = "lightblue", alpha = 0.5) +
  geom_density(aes(x = standard, color = "Standardization of IQ"), fill = "lightgreen", alpha = 0.5) +
  labs(title = "Standardized Version of IQ Distribution", x = "X-axis label", y = "Density") +
  scale_color_manual(values = c("IQ Distribution" = "blue", "Standardization of IQ" = "green")) +
  xlim(50, 165) + 
  theme_minimal()

```

::: {.callout-tip title="HG comments"}
Make sure to check your rendered document. You were missing a `+` after `xlim()` and before `theme_minimal()` which meant that it printed out the underlying function. No worries! Just worth checking. 
:::

## Question 3

*Points: 5*

::: {.callout-tip title="HG comments"}
5/5
:::

A polling firm is interested in determining how different characteristics of an individual affect vote choice. They record the following characteristics for each survey respondent - race, age, household income, and party affiliation. What "type" of variable does each characteristic most likely represent?

**Answer:**

Race - categorical variable

Age - continuous variable

Household Income - continuous Variable

Party affiliation - categorical variable

## Question 4

*Points: 5*

::: {.callout-tip title="HG comments"}
3/5
:::

Using the `states` dataset (the unit of analysis is each U.S. state), generate a box plot and density curve for each of the `vep12_turnout` and `cig_tax12` variables. Describe the distribution of each variable in detail.

::: callout-note
The `states` data set can be found in `poliscidata::states`. Take a look at `?states` to see what these variables measure.
:::

::: {.callout-tip title="HG comments"}
These are great graphs, but you needed to describe what they show. 
:::

```{r}
# poliscidata::states

vep12 <- data.frame(states$vep12_turnout)

ggplot(vep12, aes(x="", y = states$vep12_turnout)) +
  geom_boxplot(fill="pink") +
  labs(title = "Percent Turnout of Voting Eligible Population in 2012", y = "Turnout")

ggplot(vep12, aes(x=states$vep12_turnout)) +
  geom_density(alpha = 0.5, fill = "yellow") +
  labs(title = "Distribution of Turnout of the Voting Eligible Population in 2012", x = "", y = "Turnout") +
  xlim(40, 80) + 
  theme_minimal()

```

```{r}
cig_tax <- data.frame(states$cig_tax12)

ggplot(cig_tax, aes(x="", y = states$cig_tax12)) +
  geom_boxplot(fill="orange") +
  labs(title = "Cigarette Tax Per Pack in 2012", y = "Tax")

ggplot(cig_tax, aes(x=states$cig_tax12)) +
  geom_density(alpha = 10, fill = "violet") +
  labs(title = "Distribution of Cigarette Tax Per Pack in 2012", x = "", y = "Tax") +
  xlim(-1, 5) +
  theme_minimal()
```
