---
title: "Problem Set 8"
subtitle: "Due date: 20 November"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 30**

## Question 1

*Points: 5*

::: {.callout-tip}
5/5
:::

For the following regression equation, $\hat{Y} = 8.5 + 6x + \epsilon$, the standard error for $\beta_0$ is 2.5, the standard error for $\beta_1$ is 3.5, and the sample size is 2000. Find the t-statistic, 95% confidence interval, and p-value (using a two-tailed test) for $\beta_1$.

Is $\beta_1$ statistically significant at the 0.05-level with a two-tailed test? Why or why not?

```{r}
beta_1_hat <- 6
SE_beta_1 <- 3.5
sample_size1 <- 2000

t_statistic1 <- beta_1_hat / SE_beta_1

df1 <- sample_size1 - 2

critical1 <- 0.05
t_critical1 <- qt(1 - critical1 / 2, df1)

conf_interval1 <- c(beta_1_hat - t_critical1 * SE_beta_1, beta_1_hat + t_critical1 * SE_beta_1)

p_value1 <- 2 * (1 - pt(abs(t_statistic1), df1))


t_statistic1
conf_interval1
p_value1

```

**Answer - The t-statistic is 1.714286. The 95% confidence interval is -0.8640321 to 12.8640321. The p-value is 0.08663135. Since the p-value is greater than 0.05, it fails to reject the null hypothesis that** $\beta_1$ **= 0, indicating that** $\beta_1$ **is not statistically significant.**

## Question 2

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate an OLS regression and retrieve a $R^2$ value of 0.45. If the Total Sum of Squares (TSS) from that regression equals 4,700, what is the value for the Residual Sum of Squares (RSS)?

```{r}
R_squared2 <- 0.45
TSS2 <- 4700

ESS2 <- R_squared2 * TSS2

RSS2 <- TSS2 - ESS2

RSS2

```

**Answer - The value for the Residual Sum of Squares (RSS) is 2,585.**

## Question 3

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate a bivariate regression with a sample size of 102 and obtain a regression coefficient ($\beta_1$) of 5.0. What is the largest standard error that $\beta_1$ could have and still be statistically significant (i.e., reject the null hypothesis of no relationship) at the 0.05 level with a one-tailed test?

```{r}
beta_1 <- 5.0
sample_size2 <- 102
critical2 <- 0.05

df2 <- sample_size2 - 2

t_critical2 <- qt(1 - critical2, df2)

SE2_beta_1 <- beta_1 / t_critical2

SE2_beta_1
```

**Answer - The largest standard error that would still result in a statistically significant coefficient of** $\beta_1$ **at the 0.05 level with a one-tailed test is 3.011623**

## Question 4

*Points: 5*

::: {.callout-tip}
5/5
:::

Using the `states` dataset from the `poliscidata` package, produce a scatterplot of the variables `romney2012` and `hispanic10` (with `romney2012` as the dependent variable on the y-axis). Fit a regression line to the scatterplot. Describe the scatterplot and include a copy of it. Note any suspected outliers, if any (a visual inspection will suffice for this question).

::: callout-note
The variable `romney2012` measures the percentage of the state's vote that Mitt Romney received in the 2012 presidential election, and `hispanic10` indicates the percentage of the state's population that identified as Hispanic in 2010.
:::

```{r}
library(poliscidata)
library(ggplot2)

ggplot(states, aes(x = hispanic10, y = romney2012)) +
  geom_point() +            
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(x = "Population that Identified as Hispanic in 2010 (Percentage)",
       y = "Vote that Romney Received in the 2012 (Percentage)")

```

**Answer - The scattorplot indicates that as the Hispanic population increases, the percentage of states that voted for Romney decreases. Romney received the most votes when the Hispanic population was between 0 and 10 percent.**

## Question 5

*Points: 10*

::: {.callout-tip}
9/10
:::

Estimate a bivariate regression with `romney2012` as the dependent variable and `hispanic10` as the independent variable and report the results in a professionally formatted table. In as much detail as possible, describe (and interpret) the regression results.

```{r}
library(modelsummary)
library(stargazer)
# Fit bivariate regression model
regress4 <- lm(romney2012 ~ hispanic10, data = states)

regress4

stargazer(regress4, type = "text")

summary(regress4)
```

::: {.callout-tip}
Please make your coefficient labels more meaningful to your reader. 
:::

**Answer -** **The coefficient for hispanic10 is -0.196, suggesting that, on average, for each one-unit increase in hispanic10, the dependent variable romney2012 decreases by 0.196 units. The negative sign indicates a negative relationship. The coefficient for hispanic10 is not statistically significant at conventional levels of 0.1, 0.05, and 0.01 since the p-value of 0.182 is greater than all the critical values.** **The R-squared value is 0.037, indicating that the model explains 3.7% of the variance in the dependent variable. This suggests that the connection is relatively weak.**
